Live and Let Die? (Delayed Issues not Harder to Resolve)

### 2015-08-26 phaseDelay.tex
Updates from ShareLaTeX

This paper tests for the delayed issue effect in
171 software projects conducted in the period from 2006--2014 from around the world.
To the best of our knowledge,  this is the largest study
yet published on this effect.
In that sample, we found the exact
opposite of the  delayed issue effect; i.e.
the  time  to resolve 
issues  in one phase  was rarely more than when  issues
were resolved at a later phase.

@@

It is standard practice
in other fields such as medicine,
to continually revisit old conclusions~\cite{prasad13}.
Accordingly, this paper revisits
the widely-held belief of a {\em delayed issue effect} (hereafter, DIE).
Later in this paper, we offer a precise definition for this effect.
For the moment, we describe it as follows:
it is far easier to resolve issues earlier than later in the lifecycle.
 \fig{b81} shows an example of the delayed issue effect (relating
 the relative cost of fixing requirements issues at different phases of a project).

 Basili \& Boehm comment that since the 1980s,  this effect
 \begin{quote}
``...has been a major driver in focusing
industrial software practice on thorough
requirements analysis and design,
on early verification and validation, and
on up-front prototyping and simulation
to avoid costly downstream fixes''~\cite{boehm01}.
\end{quote}
 \begin{figure}[!b]  
\begin{center}
\includegraphics[angle=270,width=3in]{img/b81.pdf}
 \end{center}
 \caption{An example DIE effect. From Boehm'81~\cite{Boehm81}. }\label{fig:b81}
 \end{figure}
Other prominent authors have commented on its perceived usefulness as a rule of thumb for software engineers.  McConnell mentions it as a "common observation" in the field and  summarizes the intuitive argument for why it should be so: 
\begin{quote}
''A small mistake in upstream work can affect large amounts of downstream work. A change to a single sentence in a requirements specification can imply changes in hundreds of lines of code spread across numerous classes or modules, dozens of test cases, and numerous pages of end-user documentation''~\cite{mcconnell01}. 
\end{quote}
Glass also endorses this effect, asserting that``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.  Other researchers
are just as adamant in asserting that that the delayed issue effect is a proven
fact.
For example, what we call the delayed issued effect was listed as number one by Boehm and Basili in their "Top 10 list" of ``objective and quantitative data, relationships,
and predictive models that help
software developers avoid predictable pitfalls
and improve their ability to predict
and control efficient software projects''~\cite{boehm01}.

@@

Also, we note that our development practices have changed in ways that could   mitigate the delayed issued effect.  Previously,
software was large monolithic systems that were write once and maintain
forever. Today, there are more architectures that
support  extensive and faster changes to software. 
For example, 
Microsoft is adjusting its development practices towards a continuous
release paradigm. Theisen et al. reports their experiences where Microsoft architects
continually adjust  their modules in response to security issues~\cite{Theisen15}.
Also, upgrades to their Windows operating system is moving from service patches (which occur rarely) to continuous deployment (so there will be no Windows 11- just a stream of continuous updates to what is currently called Windows 10~\cite{bright15}).

@@

{\bf  Claim1: ``DIE'' is a  widely-held belief.}
Using a literature review, we can confirm that there are numerous historical papers (dating back decades)
that endorse DIE. Also, using a survey conducted from this paper, we find  that 
  DIE  appears up as the
single most strongly-held beliefs amongst commercial software engineers.

{\bf  Claim2: ``DIE'' is a poorly documented.}
 As discussed in our literature review,  many of the papers reporting the DIE
effect are either (1)~quite old (papers dating from last century);
(2)~that just quote prior papers without presenting   new data; 
(3)~or cite data sources that can no longer be
confirmed.

@@

That said, there are many empirical  findings that 
raise doubts that   general laws of SE even exist:
\be
\item
Turhan~\cite{me12d} lists 28 studies with contradictory conclusions
on the relation of OO measures to defects.  Those results
 directly  contradict some of the laws listed by 
Endres \& Rombach~\cite{endres03}.
\item
Ray et al.~\cite{ray2014lang} tested if   strongly typed languages
predict for better code quality. In  728 projects,
they found  only a modest benefit in strong typing
(and caution that that effect may actually be due to other  conflating factors).
\item
Fenton \& Neil~\cite{fenton00,fenton00b}   critique the truism that
``pre-release fault rates for software
are a predictor for post-release failures'' (as claimed by~\cite{dunsmore88},
amongst others). For the systems described in~\cite{fenton97}, they
show that software modules that were highly fault-prone
prior to release revealed very few faults after release.
\item
Meyer claims that   object-oriented (OO) encapsulation will
reduce error rates in software~\cite{Meyer1988}.  Yet empirical results suggest
that debugging an OO program is many times harder and
longer than debugging a standard procedural program~\cite{hatton98}.
\item
A truism of visual programming is that ``visual
representations are inherently superior to mere textual representations''. A review by Menzies suggests that the available
evidence for this claim is hardly conclusive~\cite{me00v}. 
\item
Numerous recent {\em local learning} results compare single models
learned from all available data to multiple models learned from clusters within the data~\cite{betten14,yang11,yang13,minku13,me12d,me11m,betta12,posnett11}.
A repeated result in those studies in that the local models generated the better effort
and defect predictions (better median results,
lower variance in the predictions).
\ee
% \item 
% Other work studies  found dozens of single
% models that performed worse that {\em committees of models}, all with % different opinions,
% that contribute to an {\em ensemble} prediction of software 
% development effort~\cite{kocaguneli2012value,azhar13}.
To be fair, 
SE is  not the only
field where practitioners hang on to persistent beliefs, even if the evidence
for those beliefs is not strong.
The medical profession applies  many practices based on studies
that have been disproved. For example,
a  recent article
in the Mayo Clinic Proceedings~\cite{prasad13} found  
146 medical practices based on studies 
in year $i$, but which were  reversed by subsequent trials within years $i+10$.
Even when the evidence for or against a treatment or intervention is clear, medical providers and patients may not accept it~\cite{aschwanden10}.
Aschwanden warns that ``congitive biases''  such as  confirmation bias (the tendency to look for evidence that supports what you already know and to ignore the rest)  influence how we process information~\cite{aschwanden15}.

The cognitive issues that complicate medicine are also found in software engineering.
According to Passos et al.~\cite{passos11}, commercial developers
are all too willing to believe in general
truisms.
They  warn that developers
usually assume that the truisms they learn from a few past
projects are general to 
all their future projects. They comment ``past experiences were taken into account without 
much consideration for their context~\cite{passos11}.  
The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} concur with Passos et al. They report that 
  supposed software engineering    ``experts'' rarely use lessons
  from past projects to improve their future reasoning~\cite{jorgensen09}. 
 They note that
when the experts
  fail to revise their beliefs, this leads to poor
 conclusions and software projects  (see examples in~\cite{jorgensen09}).
%% This leads to  detrimental
%% situations where groups or 
%% individuals working on the same project have conflicting beliefs, which they never debate
%or refine or unify.

@@

\section{``DIE'' is a  widely-held belief}
To assess the prevalence of DIE,   we conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
understanding of SE theory, then the DIE  may well inappropriately influence their decisions.

@@

\subsubsection{Comment 1}
Figure~\ref{fig:fix-phase-dist}
 data does not support   
{\em  requirements errors are the most expensive to fix}.
Note that   median time spent fixing errors from {\em requirements} across the life cycle was 92, 93,  49, 33, 39, 49, 67, 122 minutes\footnote{If this time seems long, then recall
that it includes the time required  to (a)~collect data and realize there is an error;
(b)~prepare a fix;  and (c)~apply some validation
procedure to check the fix (e.g. discuss it with a colleague or execute some tests).}.
This is:
\bi
\item
Clearly longer than median times seen {\em Before } and in {\em Planning}
(30,28,45,77,65,56,24,35,33);
\item
Somewhat shorter, but still not exponentially smaller,  than the median times see after {\em requirements} in {\em design}
(103,101, 79,66, 119,88, 80,45) and in {\em code} (65,115, 113,125, 95,133). 
\ei
For completeness, we ran an  
A12 test comparing the raw data (and not just the medians of \fig{raw}) and found
 a non-small effect only for the difference {\em requirements}
to {\em code}. But even though the effect was non-small, 
 the overall size of the difference
{\em requirements} to {\em code} was less than 30\% (i.e. much less than we would
expect from the delayed issue effect).

That said, Figure~\ref{fig:fix-phase-dist} does support the 
claim that early lifecycle errors are most expensive to fix.
In all our results, the biggest scale ups are found with resolving
issues introduced in the {\em before} phase. For example,
resolving {\em before} issues during unit testing was 2.56 times
slower than resolving issues earlier in the lifecycle.

Nevertheless,  it must be added that the scale up seen in 
Figure~\ref{fig:fix-phase-dist} 
is much less
than suggested by the DIE literature. Returning to
the {\em before/unit test} result: 2.56 is far smaller
than the scale ups seen in Figure~1. In fact, nowhere in 
our results do we see any exponential increase.

@@

That evidence is as follows.
Prior reports of an  exponential increase in difficulty in resolving issues {\em during
development}.
For example:
\bi
\item  From requirements to acceptance testing
\fig{b81} reports a 40-fold increase;
\item
 \fig{cost-to-fix} reports a 100-fold increase
(for the larger projects).
\item
The projects surveyed in \fig{steck} report changes of a  \newline
\{37,40,40,50,51,70,75,250\}-fold increase in pre-deployment period when the software was being developed.
\ei
Any manager watching  these intra-development trends would be well-justified
to anticipate that  the difficulty in resolving issues would will get much worse. Such managers
would therefore expect an exponentially harder time resolving issues during deployment.

On the other hand, a remarkable feature of Figure~\ref{fig:fix-phase-dist} is the {\em absence}
of an exponential delay effect during development.
In our data, the greatest difficulty in resolving requirements issues was 1.31 seen in systems testing--
which is far smaller than the 37 to 250-fold increases reported in \fig{b81}, \fig{cost-to-fix},
and \fig{steck}.  That is, from our  data, we would not anticipate exponentially greater difficulty in resolving requirements issues after
deployment. 
 \subsubsection{Comment 3}
 The above specific statements related just to requirements. Turning now
 to issues creased in all phases in Figure~\ref{fig:fix-phase-dist},  we see that
 this data does not follow the exponential increase predicted
by the delayed issue effect:
\bi
\item There is much ``up and down'' in the chart. 
\item If we separate out just the {\em Inspect} or {\em Review} data, we can see that there is some increase in {\em Design} and {\em Code}. However, that effect could be a linear or be just noise. In either case,
it does not look like exponential.
\item The time in final testing is particularly low suggesting that few defects survived into testing
(and hence, there were few instances where code required massive rework).
\ei
In summary:
\bi
\item
The delayed issue effect has been reported in around a dozen papers, at least half of which date back to the 1970s. 
\item
But there is no trace of that  effect   in hundreds of software projects built  since 2006.
\ei
%For illustrative purposes, the relative cost to fix errors per lifecycle phase in the TSP project data is overlayed with Boehm's cost curve~\cite{Boehm81} as well as the CCPDS-R case study~\cite{Royce98} in \fig{cost-to-fix-tsp}. Cost-to-fix for TSP is total time spent in each review, inspection, or test-related phase / \# of defects found in phase (reported in \fig{raw}). It is clear to see that the trend is relatively flat, closer to the pattern exhibited by the Royce case study (which also focused on early defect removal) than the exponential growth curve.

@@

\subsection{External validity}
Threats to external validity affect the generalization of findings. The most obvious threat to external validity is that TSP is not representative of most software development processes. Hence,
For example, 
one way to discount the above is that our sample is biased-- that while general software
systems suffer from the delayed issue effect, systems built using TSP are so much better controlled
that they  mitigate against the DIE effect.

It would be convenient to offer the last paragraph as the conclusion of this study
(two of our authors, Nichols and Shull,  work at the organization that invented and promoted TSP).
 However, we cannot endorse such a conclusion:
 \bi
 \item
We view TSP as a better way to
 {\em monitor} the activities {\em of  existing projects}. That is, TSP
 does not radically change a project, it just offers a better way to log the activity within
 that project. 
 \item
 Our 171 project contain examples of a wide variety
 of systems (e-commerce web portals and  banking systems) run in a variety of
 ways (agile or  waterfall or some combination of the two). Hence, rather
 than being unduly  biased in some manner,
 we view our 171 projects as a large sample of current industrial processes for software
 development.
 \ei
Another way to discount this result is to say we are only reporting on small to medium-size
projects,
(see
\fig{dist}:  median
effort was was  
271 days). 
There certainly exists another class of very large software project where early requirements
errors would be catastrophic for later development. For example, when NASA launches deep space missions,
each of those satellites are special-purpose, uniquely-built devices built by a large contractor
population. Such systems are very hard to reconfigure halfway through the development.

We concede that for very large complex NASA-like developments, the delayed issue
effect could still hold. However, perhaps we need to split our software engineering principles
into two parts- one for   very large NASA-like projects (which are not very numerous)
and one for smaller-sized projects like the ones studied here (which are far more numerous).
Most software is {\em not} like the software used on NASA deep-space satellites.
As  agile methods grew more popular last decade; and as personnel computers grew more powerful;
and as the libraries associated with particular languages grew more extensive; then smaller
teams starting  deliver systems that (previously) had required very large teams.
As mentioned above,
our  171 projects represents $\frac{171}{212}=80$\% of the software projects
seen by the Software Engineering Institute.
That is, as far as we can tell,
there are more small to medium-sized projects than otherwise.

@@

Our definition of {\em difficult to resolve} combines two concepts: time to change and cost to
change. In the above we have used them interchangable, comparing our time to resolve data (from the 171 TSP projects) against the cost-to-fix results of Boehm et al. e.g. \fig{b81}.
Is it valid to assume the equivalence of time and cost?

@@

Also, there is the issue of the implication of  a defect. Evidence discussed in \cite{Shull02} suggests that low severity defects may exhibit a lower cost to change. Nonetheless, even ``small'' typographical errors have been known to cause enormous damage (e.g., the Mars climate Orbiter).

Having documented the above issues, we assert that they are probably not major issues in the study.
As to the first point, that some debugging may be more expensive than others, one of us (Nichols)
was closely associated with many of the projects in our sample. He is unaware of any frequent
use of exorbitantly expensive tools or people on these projects. As to the second point (that
some defects have an impact beyond the mere cost to fix them), we no evidence suggesting
that the net effect of this concern over all our projects is non-zero.

@@

In this paper, we explored   the papers and data related to the 
widely-believed {\em delayed issue effect} (that delaying the resolution of issues
exponentially 
increases the difficulty of completing that  resolution).
Several prominent SE researchers state this effect is some fundamental law of software engineering~\cite{boehm01,mcconnell01,boehm01,glass02},
Based on a  survey  of both researchers and practitioners, we  found that
a specific form  of this effect (requirements errors are hardest to fix) is  widely-believed
held in the community.

@@

Our explanation for this lack-of-effect is five-fold:
\be
\item The effect might be an historical relic. Evidence:
the effect was first described in the era of punch card computing and non-interactive environments.
\item The effect might have been intermittent (rather than some fundamental law of software). Evidence: we can  found nearly
as many papers reporting the effect~\cite{Boehm76,Boehm81,steck04,Fagan76,Stephenson76} as otherwise~\cite{Royce98,Boehm80,Shull02}.
\item The effect might be confined to very large systems- in which case it would be
acceptable during development to let smaller to medium
sized projects carry some unresolved issues from early phases into later phases.
\item The effect might be mitigated by modern software development approaches that
encourage change and revision of older parts of the system.
\item The effect might be mitigated by modern software development tools
that simplify the process of large-scale reorganization of software systems.
\ee
 %One hypothesized explanation that can be ruled out is that the lack of a phase delay effect is due to using the TSP method. 
Our results beg the question: why did an idea, with so little support, become so widespread in the software engineering literature? No doubt the original evidence was compelling at the time, but much has changed in the realm of software development in the subsequent 40 years. Possibly the concept of the delayed issue effect (or its more specific description: requirements errors are the hardest to fix)
has persisted because, to use Glass's terms on the subject, it seems to be ``just common sense''\cite{glass02}. 
Nevertheless, in a rapidly changing field such as software engineering, even widely-held rules of thumb must be periodically re-verified. 
Progress in the domain of software analytics has made such periodic checks more cost-effective and feasible, and we argue that an examination of local behaviors (rather than simply accepting global heuristics) can be of significant benefit.

>>>>

This paper tests for the delayed issue effect in
171 software projects conducted around the world in the period from 2006--2014.
To the best of our knowledge,  this is the largest study
yet published on this effect.
In our sample, we found no evidence for the  delayed issue effect; i.e.
the  time  to resolve 
issues in a later phase was not consistently  more than  
when  issues were resolved soon after their introduction.

@@

It is standard practice
in other fields, such as medicine,
to continually revisit old conclusions~\cite{prasad13}.
Accordingly, this paper revisits
the widely-held belief of a {\em delayed issue effect} (hereafter, DIE).
Later in this paper, we offer a precise definition for this effect.
For the moment, we describe it as follows:
it is far easier to resolve issues earlier rather than later in the lifecycle.
 \fig{b81} shows an example of the delayed issue effect (relating
 the relative cost of fixing requirements issues at different phases of a project).

 Basili \& Boehm comment that since the 1980s,  this effect
 \begin{quote}
``...has been a major driver in focusing
industrial software practice on thorough
requirements analysis and design,
on early verification and validation, and
on up-front prototyping and simulation
to avoid costly downstream fixes''~\cite{boehm01}.
\end{quote}
 \begin{figure}[!b]  
\begin{center}
\includegraphics[angle=270,width=3in]{img/b81.pdf}
 \end{center}
 \caption{An example DIE effect. From Boehm'81~\cite{Boehm81}. }\label{fig:b81}
 \end{figure}
Other prominent authors have commented on its perceived usefulness as a rule of thumb for software engineers.  McConnell mentions it as a "common observation" in the field and  summarizes the intuitive argument for why it should be so: 
\begin{quote}
''A small mistake in upstream work can affect large amounts of downstream work. A change to a single sentence in a requirements specification can imply changes in hundreds of lines of code spread across numerous classes or modules, dozens of test cases, and numerous pages of end-user documentation''~\cite{mcconnell01}. 
\end{quote}
Glass also endorses this effect, asserting that ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.  Other researchers
are just as adamant in asserting that that the delayed issue effect is a proven
fact.
For example, what we call the delayed issued effect was listed first by Boehm and Basili in their "Top 10 list" of ``objective and quantitative data, relationships,
and predictive models that help
software developers avoid predictable pitfalls
and improve their ability to predict
and control efficient software projects''~\cite{boehm01}.

@@

Also, we note that our development practices have changed in ways that could   mitigate the delayed issued effect.  Previously,
software was large monolithic systems that were ''write once and maintain
forever.'' Today, there are more architectures that
support  extensive and faster changes to software. 
For example, 
Microsoft is adjusting its development practices towards a continuous
release paradigm. Theisen et al. reports their experiences where Microsoft architects
continually adjust  their modules in response to security issues~\cite{Theisen15}.
Also, upgrades to their Windows operating system is moving from service patches (which occur rarely) to continuous deployment (so there will be no Windows 11- just a stream of continuous updates to what is currently called Windows 10~\cite{bright15}).

@@

{\bf  Claim1: ``DIE'' is a  widely-held belief.}
Using a literature review, we can confirm that there are numerous historical papers (dating back decades)
that endorse DIE. Also, using a survey conducted from this paper, we find  that 
  DIE  appears as the
single most strongly-held beliefs amongst commercial software engineers.

{\bf  Claim2: ``DIE'' is a poorly documented.}
 As discussed in our literature review,  many of the papers reporting the DIE
effect are either (1)~quite old (papers dating from last century);
(2)~quoting prior papers without presenting   new data; 
(3)~or citing data sources that can no longer be
confirmed.

@@

That said, there are many empirical  findings that 
raise doubts that   general laws of SE even exist:
\be
\item
Turhan~\cite{me12d} lists 28 studies with contradictory conclusions
on the relation of OO measures to defects.  Those results
 directly  contradict some of the laws listed by 
Endres \& Rombach~\cite{endres03}.
\item
Ray et al.~\cite{ray2014lang} tested if   strongly typed languages
predict for better code quality. In  728 projects,
they found  only a modest benefit in strong typing
(and caution that that effect may actually be due to other  conflating factors).
\item
Fenton \& Neil~\cite{fenton00,fenton00b}   critique the truism that
``pre-release fault rates for software
are a predictor for post-release failures'' (as claimed by~\cite{dunsmore88},
amongst others). For the systems described in~\cite{fenton97}, they
show that software modules that were highly fault-prone
prior to release revealed very few faults after release.
\item
Meyer claims that   object-oriented (OO) encapsulation will
reduce error rates in software~\cite{Meyer1988}.  Yet empirical results suggest
that debugging an OO program is many times harder and
longer than debugging a standard procedural program~\cite{hatton98}.
\item
A truism of visual programming is that ``visual
representations are inherently superior to mere textual representations''. A review by Menzies suggests that the available
evidence for this claim is hardly conclusive~\cite{me00v}. 
\item
Numerous recent {\em local learning} results compare single models
learned from all available data to multiple models learned from clusters within the data~\cite{betten14,yang11,yang13,minku13,me12d,me11m,betta12,posnett11}.
A repeated result in those studies is that the local models generated the better effort
and defect predictions (better median results,
lower variance in the predictions).
\ee
% \item 
% Other work studies  found dozens of single
% models that performed worse that {\em committees of models}, all with % different opinions,
% that contribute to an {\em ensemble} prediction of software 
% development effort~\cite{kocaguneli2012value,azhar13}.
To be fair, 
SE is  not the only
field where practitioners hang on to persistent beliefs, even if the evidence
for those beliefs is not strong.
The medical profession applies  many practices based on studies
that have been disproved. For example,
a  recent article
in the Mayo Clinic Proceedings~\cite{prasad13} found  
146 medical practices based on studies 
in year $i$, but which were  reversed by subsequent trials within years $i+10$.
Even when the evidence for or against a treatment or intervention is clear, medical providers and patients may not accept it~\cite{aschwanden10}.
Aschwanden warns that ``cognitive biases''  such as  confirmation bias (the tendency to look for evidence that supports what you already know and to ignore the rest)  influence how we process information~\cite{aschwanden15}.

The cognitive issues that complicate medicine are also found in software engineering.
According to Passos et al.~\cite{passos11}, commercial developers
are all too willing to believe in general
truisms.
They  warn that developers
usually assume that the truisms they learn from a few past
projects are general to 
all their future projects. They comment ``past experiences were taken into account without 
much consideration for their context~\cite{passos11}.''  
The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} concur with Passos et al. They report that 
  supposed software engineering    ``experts'' rarely use lessons
  from past projects to improve their future reasoning~\cite{jorgensen09}. 
 They note that
when the experts
  fail to revise their beliefs, this leads to poor
 conclusions and software projects  (see examples in~\cite{jorgensen09}).
%% This leads to  detrimental
%% situations where groups or 
%% individuals working on the same project have conflicting beliefs, which they never debate
%or refine or unify.

@@

\section{``DIE'' is a  widely-held belief}
To assess the prevalence of DIE,   we conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
understanding of SE theory, then the DIE  may well influence their decisions.

@@

\subsubsection{Comment 1}
Figure~\ref{fig:raw}
 data does not support that   
{\em  requirements errors are the most expensive to fix}.
Note that   median time spent fixing errors from {\em requirements} across the life cycle was 92, 93,  49, 33, 39, 49, 67, 122 minutes\footnote{If this time seems long, then recall
that it includes the time required  to (a)~collect data and realize there is an error;
(b)~prepare a fix;  and (c)~apply some validation
procedure to check the fix (e.g. discuss it with a colleague or execute some tests).}.
This is:
\bi
\item
Clearly longer than median times seen {\em Before } and in {\em Planning}
(30,28,45,77,65,56,24,35,33);
\item
Somewhat shorter, but still not exponentially smaller,  than the median times see after {\em requirements} in {\em design}
(103,101, 79,66, 119,88, 80,45) and in {\em code} (65,115, 113,125, 95,133). 
\ei
For completeness, we ran an A12 test comparing the raw data for each phase, and found an effect size of ''medium'' or higher only for the difference between {\em requirements}
and {\em code}. In this case, however, 
 the overall size of the difference
{\em requirements} to {\em code} was less than 30\% (i.e. much less than we would
expect from the delayed issue effect).

That said, Figure~\ref{fig:raw} does support the 
claim that early lifecycle errors are most expensive to fix.
In all our results, the biggest scale ups are found with resolving
issues introduced in the {\em before} phase. For example,
resolving {\em before} issues during unit testing was 2.56 times
slower than resolving issues earlier in the lifecycle.

Nevertheless, the scale up seen in 
Figure~\ref{fig:raw} 
is much less
than that suggested by the DIE literature. Returning to
the {\em before/unit test} result: 2.56 is far smaller
than the scale ups seen in Figure~1. In fact, nowhere in 
our results do we see any exponential increase.

@@

That evidence is as follows.
Prior literature reports an  exponential increase in difficulty in resolving issues {\em during
development}.
For example:
\bi
\item  From requirements to acceptance testing
\fig{b81} reports a 40-fold increase;
\item
 \fig{cost-to-fix} reports a 100-fold increase
(for the larger projects).
\item
The projects surveyed in \fig{steck} report changes of a  \newline
\{37,40,40,50,51,70,75,250\}-fold increase in pre-deployment period when the software was being developed.
\ei
Any manager watching  these intra-development trends would be well-justified
to anticipate that  the difficulty in resolving issues  will get much worse. Such managers
would therefore expect an exponentially harder time resolving issues during deployment.

On the other hand, a remarkable feature of Figure~\ref{fig:raw} is the {\em absence}
of an exponential delay effect during development.
In our data, the greatest increase in difficulty in resolving requirements issues was the 1.31 scale-up seen in systems testing--
which is far smaller than the 37 to 250-fold increases reported in \fig{b81}, \fig{cost-to-fix},
and \fig{steck}.  That is, from our  data, we would not anticipate exponentially greater difficulty in resolving requirements issues after
deployment. 
 \subsubsection{Comment 3}
 The above specific statements are related just to requirements. Turning now
 to issues created in all phases in Figure~\ref{fig:raw},  we see that
 this data does not follow the exponential increase predicted
by the delayed issue effect:
\bi
\item There is much ``up and down'' in the chart. 
\item If we separate out just the {\em Inspect} or {\em Review} data, we can see that there is some increase in {\em Design} and {\em Code}. However, that effect could be a linear or be just noise. In either case,
it does not look like exponential.
\item The time in final testing is particularly low suggesting that few defects survived into testing
(and hence, there were few instances where code required massive rework).
\ei
%In summary:
%\bi
%\item
%The delayed issue effect has been reported in around a dozen papers, at least half of which date back to the 1970s. 
%\item
%But there is no trace of that  effect   in hundreds of software projects built  since 2006.
%\ei
%For illustrative purposes, the relative cost to fix errors per lifecycle phase in the TSP project data is overlayed with Boehm's cost curve~\cite{Boehm81} as well as the CCPDS-R case study~\cite{Royce98} in \fig{cost-to-fix-tsp}. Cost-to-fix for TSP is total time spent in each review, inspection, or test-related phase / \# of defects found in phase (reported in \fig{raw}). It is clear to see that the trend is relatively flat, closer to the pattern exhibited by the Royce case study (which also focused on early defect removal) than the exponential growth curve.

@@

\subsection{External validity}
Threats to external validity affect the generalization of findings. The most obvious threat to external validity is that TSP is not representative of most software development processes. Hence,
for example, 
one way to discount the above argumentation is that our sample is biased-- that while general software
systems suffer from the delayed issue effect, systems built using TSP are so much better controlled
that they  mitigate against the DIE effect.

It would be convenient to offer the last paragraph as the conclusion of this study
(two of our authors, Nichols and Shull,  work at the organization that invented and promoted TSP).
 However, we cannot endorse such a conclusion:
 \bi
 \item
We view TSP as a better way to
 {\em monitor} the activities {\em of  existing projects}. That is, TSP
 does not radically change a project, it just offers a better way to log the activity within
 that project. 
 \item
 Our 171 projects contain examples of a wide variety
 of systems (ranging from e-commerce web portals to  banking systems) run in a variety of
 ways (agile or  waterfall or some combination of the two). Hence, rather
 than being unduly  biased in some manner,
 we view our 171 projects as a large sample of current industrial processes for software
 development.
 \ei
Another way to discount this result is to say we are only reporting on small to medium-size
projects
(see
\fig{dist}:  median
effort was was  
271 days). 
There certainly exists another class of very large software project where early requirements
errors would be catastrophic for later development. For example, when NASA launches deep space missions,
each of those satellites are special-purpose, uniquely-built devices built by a large contractor
population. Such systems are very hard to reconfigure halfway through the development.

We concede that for very large complex NASA-like developments, the delayed issue
effect could still hold. However, perhaps we need to split our software engineering principles
into two parts- one for   very large NASA-like projects (which are not very numerous)
and one for smaller-sized projects like the ones studied here (which are far more numerous).
Most software is {\em not} like the software used on NASA deep-space satellites.
As  agile methods grew more popular last decade; and as personal computers grew more powerful;
and as the libraries associated with particular languages grew more extensive; then smaller
teams starting  deliver systems that (previously) had required very large teams.
As mentioned above,
our  171 projects represents $\frac{171}{212}=80$\% of the software projects
seen by the Software Engineering Institute's TSP team.
That is, as far as we can tell,
there are more small to medium-sized projects than otherwise.

@@

Our definition of {\em difficult to resolve} combines two concepts: time to change and cost to
change. In the above we have used them interchangably, comparing our time to resolve data (from the 171 TSP projects) against the cost-to-fix results of Boehm et al. e.g. \fig{b81}.
Is it valid to assume the equivalence of time and cost?

@@

Also, there is the issue of the implication of  a defect. Evidence discussed in \cite{Shull02} suggests that low severity defects may exhibit a lower cost to change. Nonetheless, even ``small'' errors have been known to cause enormous damage (e.g., the Mars Climate Orbiter).

Having documented the above issues, we assert that they are unlikely to be major issues in the study.
As to the first point, that some debugging may be more expensive than others, one of us (Nichols)
was closely associated with many of the projects in our sample. He is unaware of any frequent
use of exorbitantly expensive tools or people on these projects. As to the second point (that
some defects have an impact beyond the mere cost to fix them), we have no evidence suggesting
that the net effect of this concern over all our projects is non-zero.

@@

In this paper, we explored   the papers and data related to the 
widely-believed {\em delayed issue effect} (that delaying the resolution of issues
exponentially 
increases the difficulty of completing that  resolution).
Several prominent SE researchers state this effect is a fundamental law of software engineering~\cite{boehm01,mcconnell01,boehm01,glass02}.
Based on a  survey  of both researchers and practitioners, we  found that
a specific form  of this effect (requirements errors are hardest to fix) is  widely-believed
held in the community.

@@

To be clear, we do not claim that this effect {\em never} holds in software projects; just that it cannot be assumed to {\em always} hold. Our explanation of the observed lack-of-effect is five-fold:
\be
\item The effect might be an historical relic. Evidence:
the effect was first described in the era of punch card computing and non-interactive environments.
\item The effect might have been intermittent (rather than some fundamental law of software). Evidence: we can  found nearly
as many papers reporting the effect~\cite{Boehm76,Boehm81,steck04,Fagan76,Stephenson76} as otherwise~\cite{Royce98,Boehm80,Shull02}.
\item The effect might be confined to very large systems- in which case it would be
acceptable during development to let smaller to medium
sized projects carry some unresolved issues from early phases into later phases.
\item The effect might be mitigated by modern software development approaches that
encourage change and revision of older parts of the system.
\item The effect might be mitigated by modern software development tools
that simplify the process of large-scale reorganization of software systems.
\ee
 %One hypothesized explanation that can be ruled out is that the lack of a phase delay effect is due to using the TSP method. 
Our results beg the question: why did an idea, with so little support, become so widespread in the software engineering literature? No doubt the original evidence was compelling at the time, but much has changed in the realm of software development in the subsequent 40 years. Possibly the concept of the delayed issue effect (or its more specific description: requirements errors are the hardest to fix)
has persisted because, to use Glass's terms on the subject, it seems to be ``just common sense''\cite{glass02}. 
Nevertheless, in a rapidly changing field such as software engineering, even widely-held rules of thumb must be periodically re-verified. 
Progress in the domain of software analytics has made such periodic checks more cost-effective and feasible, and we argue that an examination of local behaviors (rather than simply accepting global heuristics) can be of significant benefit.
